Cocoa: Congestion Control Aware Queuing

### 2019-11-28 tex/paper.tex
Update presentation

% Modify this to remove all ACM and conference information (for arXiv for example)
\def\removeHeaders{no}

\def\tempYes{yes}
\def\tempNo{no}
\documentclass[10pt,sigconf,letterpaper,dvipsnames\ifx\removeHeaders\tempYes ,nonacm\fi]{acmart}
\acmYear{2019}
\copyrightyear{2019}
\acmConference[BS'19]{Buffer Sizing Workshop}{December 2-3}{Stanford, CA}

@@

\ifx\removeHeaders\tempYes
\settopmatter{printacmref=false} % Removes citation information below abstract
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
\fi

@@

\begin{abstract}
Recent model-based congestion control algorithms such as BBR use repeated measurements at the endpoint to build a model of the network connection and use it to achieve optimal throughput with low queuing delay. Conversely, applying this model-based approach to Active Queue Management (AQM) has so far received less attention. We propose the new AQM scheduler \textit{cocoa} based on fair queuing, which adapts the buffer size depending on the needs of each flow without requiring active participation from the endpoint. We implement this scheduler for the Linux kernel and show that it interacts well with the most common congestion control algorithms and can significantly increase throughput compared to \codel{} while avoiding overbuffering.
\end{abstract}

@@

We show that a prototype of our solution \textit{\gls{cocoa}} can achieve these objectives for the most common loss-based \glspl{cca}, Reno \cite {jacobson_congestion_1988} and Cubic \cite{ha_cubic:_2008}. These \glspl{cca} operate by continuously increasing the number of bytes that are allowed to be in the network (congestion window) if no packet loss is experienced and by sharply decreasing this number if a packet is lost (multiplicative decrease). With Cubic being the default \gls{cc} in all major OS, we especially emphasize our evaluation on improving its performance. Contrasting to the aforementioned \glspl{cca}, recently proposed BBR \cite{cardwell_bbr:_2016} does not continuously increase and then sharply decrease when packets are lost but instead uses periodic measurements to estimate the available bandwidth as well as the minimal \gls{rtt} and then tries to stay at this point of optimal bandwidth and minimal delay. BBR is thus considered to be \textit{model-based}. We show that \gls{cocoa} also behaves well in interaction with BBR~v1.

@@

The purpose of the multiplier, which is used to multiply the \gls{li} of previous \gls{gi} to define the new \gls{gi}, is to prevent that the short gap between two subsequent packet losses is used as the interval to determine whether the buffer size has to be adapted. This means that the multiplier must be large enough so that it allows to skip over the packet losses that occur when the buffer is full. For Reno and Cubic, the loss-free period is very large compared to the period during which packet loss occurs. Specifically, the period with packet loss has the length of one \gls{rtt} since Reno and Cubic reduce their sending rate as soon as packet loss occurs, thus ending the packet loss. This means that the multiplier can be very small in order to work for these \glspl{cca}. We conducted experiments with the multiplier being $0.5$ and have never encountered the problem that the interval between adjacent packet losses is considered the longest interval and that the buffer is cut to virtually nothing because a standing queue is assumed. However, BBR~v1 uses a completely different approach compared to Reno and Cubic to probe for bandwidth: It periodically increases its sending rate and then reduces it again to see if it can achieve more throughput. This happens with the following pattern: $[\frac{5}{4},\frac{3}{4},1,1,1,1,1,1]$. The sending rate is increased to probe for more bandwidth, then decreased to reduce the standing queue potentially formed and after that BBR is simply keeping the sending rate to match the estimated bandwidth. Each of these eight phases lasts one \gls{rtt}. BBR uses a random cyclic permutation of this pattern with the only condition being that it cannot start with $\frac{3}{4}$. The following pattern could thus occur: $[1,1,1,1,1,1,\frac{5}{4},\frac{3}{4},\frac{5}{4},\frac{3}{4},1,1,1,1,1,1]$. If the \gls{gi} ends after the first 6 phases and also the \gls{li} was these six phase, with a factor of $0.5$, the next \gls{gi} would thus end at the end of phase 8. Then, the \gls{li} would be the 8th cycle. The next \gls{gi} would then be of the duration of $0.5$ phases and would then end in the middle of cycle 9. This would be erroneous because this is a probing phase during which packet loss occurs. The aforementioned problem of erroneously and significantly reducing the buffer could thus occur. This means that for BBR~v1 the multiplier must be larger than 1 since it can happen that non-probing phases are followed by probing phases of the same length. From these considerations it follows that for our experiments we generally use a multiplier of $1.25$. The behavior is different for BBR~v2 which drastically changed the probing phase and now consists of a random 2-3\,s phase of constant sending rate (cruising) followed by a quick probing phase followed by a decrease phase \cite{cardwell_bbr_2019}. We thus conjecture that a multiplier of $0.5$ is sufficient for BBR~v2. However, we could not finally confirm this since BBR~v2 is still under development.

@@

First we evaluate if \gls{cocoa} is able to maintain a small buffer like it is necessary in case of small \glspl{bdp}. As can be seen in \autoref{fig:fq} the regular fq qdisc maintains a standing queue and unnecessarily leads to a significant increase in \gls{rtt} (100\,ms minimum, when the real minimum is 10\,ms). In contrast, \autoref{fig:cn} maintains the full throughput while at the same time keeping the delay minimal and not having a standing queue. Also, when drastically changing the bandwidth by halving it (\autoref{fig:cn}) or doubling it (\autoref{fig:cn2}) \gls{cocoa} rapidly adapts to the new conditions and returns to the optimum state.

@@

Besides fq\_codel possibly leading to lower throughput as we have shown, another problem is that for links with a small \gls{bdp} fq\_codel can keep a standing queue akin to fq since fq\_codel only wants queuing delay to fall below 5\,ms once every 100\,ms. %For example, on a link with a base \gls{rtt} of 5\,ms, fq\_codel would accept a permanent standing queue of 4.9\,ms, leading to overall \gls{rtt} being doubled, which is not required for achieving optimal throughput. 
We also ran experiments for these low \gls{bdp} scenarios: For a link speed of 20\;Mbit/s and a base \gls{rtt} of 1\;ms for a Cubic flow both \gls{cocoa} and fq\_codel achieve 98\% utilization. The average \gls{rtt} is 6.86\,ms for \gls{cocoa} and 7.31\;ms for fq\_codel.

@@

We see this work as an initial step towards flow-adaptive fair queuing at bottlenecks. An interesting further improvement could be to not statically initialize the buffer with a constant 100 packets but instead use experience from previous flows. Moreover, we think that a promising direction for future work might be to explore the use of reinforcement learning. Such an approach would fingerprint a flow and choose the buffer size accordingly, maximizing a chosen objective such as high throughput with minimal delay.

>>>>

% Modify this to remove all ACM and conference information (for arXiv for example)
\def\removeHeaders{yes}

\def\tempYes{yes}
\def\tempNo{no}
\documentclass[sigconf, 10pt, letterpaper,dvipsnames\ifx\removeHeaders\tempYes ,nonacm\fi]{acmart}

%\documentclass[sigconf]{acmart}

% Copyright
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference info
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

\settopmatter{printacmref=false, printccs=false, printfolios=true}

% DOI
\acmDOI{}

% ISBN
\acmISBN{}

%Conference
%\acmConference[Submitted for review to SIGCOMM]{}
%\acmYear{2018}
%\copyrightyear{}

%% {} with no args suppresses printing of the price
\acmPrice{}

\acmYear{2019}
\copyrightyear{2019}
\acmConference[BS'19]{Buffer Sizing Workshop}{December 2-3}{Stanford, CA}

@@

%\ifx\removeHeaders\tempYes
%\settopmatter{printacmref=false} % Removes citation information below abstract
%\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
%\fi

@@

\begin{abstract}
Recent model-based congestion control algorithms such as BBR use repeated measurements at the endpoint to build a model of the network connection and use it to achieve optimal throughput with low queuing delay. Conversely, applying this model-based approach to Active Queue Management (AQM) has so far received less attention. We propose the new AQM scheduler \textit{cocoa} based on fair queuing, which adapts the buffer size depending on the needs of each flow without requiring active participation from the endpoint. We implement this scheduler for the Linux kernel and show that it interacts well with the most common congestion control algorithms and can significantly increase throughput compared to fair \codel{} while avoiding overbuffering.
\end{abstract}

@@

We show that a prototype of our solution \textit{\gls{cocoa}} can achieve the aforementioned objectives for the most common loss-based \glspl{cca}, Reno \cite {jacobson_congestion_1988} and Cubic \cite{ha_cubic:_2008}. These \glspl{cca} operate by continuously increasing the number of bytes that are allowed to be in the network (congestion window) if no packet loss is experienced and by sharply decreasing this number if a packet is lost (multiplicative decrease). With Cubic being the default \gls{cc} in all major OS, we especially emphasize our evaluation on improving its performance. Contrasting to the aforementioned \glspl{cca}, recently proposed BBR \cite{cardwell_bbr:_2016} does not continuously increase and then sharply decrease when packets are lost but instead uses periodic measurements to estimate the available bandwidth as well as the minimal \gls{rtt} and then tries to stay at this point of optimal bandwidth and minimal delay. BBR is thus considered to be \textit{model-based}. We show that \gls{cocoa} also behaves well in interaction with BBR~v1.

@@

The purpose of the multiplier, which is used to multiply the \gls{li} of previous \gls{gi} to define the new \gls{gi}, is to prevent that the short gap between two subsequent packet losses is used as the interval to determine whether the buffer size has to be adapted. This means that the multiplier must be large enough so that it allows to skip over the packet losses that occur when the buffer is full. For Reno and Cubic, the loss-free period is very large compared to the period during which packet loss occurs. Specifically, the period with packet loss has the length of one \gls{rtt} since Reno and Cubic reduce their sending rate as soon as packet loss occurs, thus ending the packet loss. This means that the multiplier can be very small in order to work for these \glspl{cca}. We conducted experiments with the multiplier being $0.5$ and have never encountered the problem that the interval between adjacent packet losses is considered the longest interval and that the buffer is cut to virtually nothing because a standing queue is assumed. However, BBR~v1 uses a completely different approach compared to Reno and Cubic to probe for bandwidth: It periodically increases its sending rate and then reduces it again to see if it can achieve more throughput. This happens with the following pattern: $[\frac{5}{4},\frac{3}{4},1,1,1,1,1,1]$. The sending rate is increased to probe for more bandwidth, then decreased to reduce the standing queue potentially formed and after that BBR is simply keeping the sending rate to match the estimated bandwidth. Each of these eight phases lasts one \gls{rtt}. BBR uses a random cyclic permutation of this pattern with the only condition being that it cannot start with $\frac{3}{4}$. The following pattern could thus occur: $[1,1,1,1,1,1,\frac{5}{4},\frac{3}{4},\frac{5}{4},\frac{3}{4},1,1,1,1,1,1]$. If the \gls{gi} ends after the first 6 phases and also the \gls{li} was these six phase, with a factor of $0.5$, the next \gls{gi} would thus end at the end of phase 8. Then, the \gls{li} would be the 8th cycle. The next \gls{gi} would then be of the duration of $0.5$ phases and would then end in the middle of cycle 9. This would be erroneous because this is a probing phase during which packet loss occurs. The aforementioned problem of erroneously and significantly reducing the buffer could thus occur. This means that for BBR~v1 the multiplier must be larger than 1 since it can happen that non-probing phases are followed by probing phases of the same length. From these considerations it follows that for our experiments we generally use a multiplier of $1.25$. The behavior is different for BBR~v2 which drastically changed the probing phase and now consists of a random 2-3\,s phase with a constant sending rate (cruising) followed by a quick probing phase followed by a decrease phase \cite{cardwell_bbr_2019}. We thus conjecture that a multiplier of $0.5$ is sufficient for BBR~v2. However, we could not finally confirm this since BBR~v2 is still under development.

@@

First we evaluate if \gls{cocoa} is able to maintain a small buffer like it is necessary in case of small \glspl{bdp}. As can be seen in \autoref{fig:fq} the regular fq qdisc (with a standard queue size of 100 packets) maintains a standing queue and unnecessarily leads to a significant increase in \gls{rtt} (100\,ms minimum, when the real minimum is 10\,ms). In contrast, \autoref{fig:cn} maintains full throughput while at the same time keeping the delay minimal and not having a standing queue. Also, when drastically changing the bandwidth by halving it (\autoref{fig:cn}) or doubling it (\autoref{fig:cn2}) \gls{cocoa} rapidly adapts to the new conditions and returns to the optimum state.

@@

Besides fq\_codel possibly leading to lower throughput as we have shown, another problem is that for links with a small \gls{bdp} fq\_codel can keep a standing queue akin to fq since fq\_codel only wants queuing delay to fall below 5\,ms once every 100\,ms. For example, on a link with a base \gls{rtt} of 1\,ms, fq\_codel would accept a permanent standing queue of 4.9\,ms, leading to overall \gls{rtt} being more than doubled, which is not required for achieving optimal throughput. 
%We also ran experiments for these low \gls{bdp} scenarios: For a link speed of 20\;Mbit/s and a base \gls{rtt} of 1\;ms for a Cubic flow both \gls{cocoa} and fq\_codel achieve 98\% utilization. The average \gls{rtt} is 6.86\,ms for \gls{cocoa} and 7.31\;ms for fq\_codel.

@@

We see this work as an initial step towards learning and flow-adaptive fair queuing at bottlenecks and think that the most promising application domain of \gls{cocoa} is at the Internet's edge. An interesting further improvement could be to not statically initialize the buffer with a constant 100 packets but instead use experience from previous flows. Moreover, we think that a promising direction for future work might be to explore the use of reinforcement learning. Such an approach would fingerprint a flow and choose the buffer size accordingly, maximizing a chosen objective such as high throughput with minimal delay.
